A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-09-14 demystify.tex
updates
Updates from Overleaf

Most well-known and well-respected profs. would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). However, many international students do not know this (e.g., it is a common practice at their university) and might be in a situation they do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer, relatively unknown in the field, or simply do not know/care enough about you.  In any of these cases, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

@@

\begin{commentbox}[Thanh:]
  Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?
  \tcblower
  Vu: One way is looking at what well-known researchers publish. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  At Mason, faculty are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potential and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected.
  \tcblower
  In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.
\end{commentbox}

@@

  In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
  \tcblower
  \textbf{Vu}: \autoref{sec:contact} shows how to contact a professor for research opportunities. Many probably will say no (or do not reply) as they do not have the bandwidth to take on random students, but some may say yes if they see a potential fit.
\end{commentbox}

@@

\begin{commentbox}[Dear Prof. Nguyen,]

  I am writing to inquire about potential research opportunities as a GRA in your group at Mason. Currently I am an undergraduate student in Computer Science at UNIV and plan to graduate in May 2023.

@@

  This is a good example because it is written just for me.  It shows that the student knows about my work on invariant generation and has a related background (paper C and project G).
\end{commentbox}

@@

  \item [PI]  Principal Investigator is lead researcher on a grant or research project, e.g., a PI on an NSF grant.  Prospective students often use this term to refer to a professor they are interested in working with.  However, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1/R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. Research 2 (R2) universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically larger, and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.

@@

\begin{commentbox}[Vu:]
    Mason has a very large the number of MS students in CS (1000+ MS vs 200 PhDs). In contrast, other similar size universities often have much smaller MS CS programs (or none). Location plays an important role as GMU is close to DC with many developer professionals who want to get an MS, which are often covered by their employers.  Our MS program is geared towards working professionals, e.g., all of our MS courses are offered in the evening and online.

@@

\subsection{Funding}
Unlike PhD programs, which typically guarantee funding (\autoref{sec:funding}), MS programs are typically not funded. You have to pay tuition, which can be expensive. It is difficult for an MS student to get funding, e.g., through TA or RA positions, because these are typically reserved for PhD students.
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive, then a prof. might be willing to take you as an RA. Moreover, some schools do have TAs for MS students, and you can apply for these positions. For example, Mason CS dept has TAs for MS students (of course not for all of our 1000+ MS students, but quite a few); my courses in the past two years have MS TAs.

@@

\begin{commentbox}[Vu:]
    I maintain a close relationship with my former professors and mentors. When there is a new event in my life (or theirs), I often email them or call them, e.g., when I get married, have a new baby, new job, etc. When I was having issues at UNL and considering leaving, I talked to my PhD advisers Deepak and Steph. I think this does not bother them a bit; they are genuinely interested in knowing and helping solve these ``dramas'' in my life.

    I also visit my former professors when I am in their area. I meet Thang Bui (my MS adviser) at least once a year when I come back to Harrisburg to visit my parents. When Steph was in DC for a meeting, I invited her to to give a research talk at Mason. I have also collaborated with them after I graduated e.g., I am working with Deepak on an NSF proposal on invariant generation.

>>>>

Most well-known and well-respected profs. would not ask you to write your own letter (in fact, even not well-known ones wouldn't do this). This might be a common practice at their university or the students do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they are likely not a good English writer, relatively unknown in the field, or simply do not know/care enough about you.  In any of these cases, their letters won't worth much. So if you are in this situation, you should find someone else to write for you.

@@

\begin{commentbox}[Thanh:]
  Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?
  \tcblower
  \textbf{Vu}: One way is looking at what well-known researchers publish. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  At Mason, faculty are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potential and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected.\\

  In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.
\end{commentbox}

@@

  In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
  \tcblower
  \textbf{Vu}: \autoref{sec:contact} shows how to contact a professor for research opportunities. Many will say no (or do not reply) as they do not have the bandwidth to take on random students, but some may say yes if they see a potential fit.
\end{commentbox}

@@

\begin{commentbox}[Dear Prof. Nguyen,]\\

  I am writing to inquire about potential research opportunities as a GRA in your group at Mason. Currently I am an undergraduate student in Computer Science at UNIV and plan to graduate in May 2023.
    \\

@@

 \tcblower
  This is a good example because it is written just for me.  It shows that the student knows about my work on invariant generation and has a related background (paper C and project G).
\end{commentbox}

@@

  \item [PI]  Principal Investigator is lead researcher on a grant or research project, e.g., a PI on an NSF grant.  Prospective students often use this term to refer to a professor they are interested in working with.  However, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1/R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. R2 universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically larger, and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.

@@

\begin{commentbox}[Vu:]
    Mason has a very large the number of MS students in CS (1000+ MS vs 200 PhDs). In contrast, other similar size universities often have much smaller MS CS programs (or none). Location plays an important role as GMU is close to DC with many developer professionals who want to get an MS, which are often covered by their employers.  Our MS program is geared towards working professionals, e.g., all of our MS courses are offered in the evening and online.
    \\

@@

\subsection{Funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded through either RA or TA.
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive, then a prof. might be willing to take you as an RA. Moreover, some schools do have TAs for MS students, and you can apply for these positions. For example, Mason CS dept has quite a few TAs for MS students; my courses in the past two years have MS TAs.

@@

\begin{commentbox}[Vu:]
    I maintain a close relationship with my former professors and mentors. When there is a new event in my life (or theirs), I often email them or call them, e.g., when I get married, have a new baby, new job, etc. When I was having issues at UNL and considering leaving, I talked to my PhD advisers Deepak and Steph. I think this does not bother them a bit; they are genuinely interested in knowing and helping solve these ``dramas'' in my life.
    \\

    I also visit my former professors when I am in their area. I meet Thang Bui (my MS adviser) at least once a year when I come back to Harrisburg to visit my parents. When Steph was in DC for a meeting, I invited her to to give a research talk at Mason. I have also collaborated with them after I graduated e.g., I recently got an NSF grant with Deepak.
    \\

### 2024-09-14 faculty.tex
Updates from Overleaf

While teaching takes about 40\% of your time, it is not as nearly as important as research for tenure.  Moreover, for most new faculty, teaching is not as difficult as getting grants or publishing papers, and thus is often taken lightly by new faculty.  However, teaching is still important for your tenure case, and being a good teacher can be very rewarding. Moreover, innovative integration of research and teaching can help you get good students and even NSF grants.

\section{Teaching Load}
New TT faculty often have a lighter teaching load in their first several years or even throughout their tenure track.  This is to give you time to establish your research program. The teaching load is usually higher at teaching universities, and lower at research universities.

The desirable teaching load for TT at most R1 universities is 1+1 (1 course / semester). While some univs might give you a lighter load in your first year, most univs expect you to teach 1+1 from your second year. In many cases, 1+1 is only for the first 2 or 3 years and after that it increases to 1+2 (or 2+1). It is rare that a TT in CS at R1 universities would have to teach 2+2 during the tenure track.

In many cases, your contract might list 1+1 for the first X year and 1+2 after until tenure, but the dept unofficial gives TT faculty 1+1 until tenure. Some other ways to reduce teaching load such as \emph{buying out} of teaching with a grant, leading a very large grant, or have taught many large classes.

\begin{commentbox}[Vu]

    At UNL, TT faculty is 1+1 for the first 3 years, and 1+2 after that.  At GMU, it is 1+1 for the first 3 years, but I haven't seen any TT having to teach more than that. In addition, GMU grants all TT faculty a semester off from teaching before tenure. Some faculty use this to work on their NSF CAREER proposal while others use it to prepare their tenure application.

  \end{commentbox}

\section{Courses}TT faculty will teach both undergrad and grad courses.  Grad courses include seminars, special topics, and core courses.  Most new faculty use seminars and special topics to recruit students for their research group.  Undergrad courses are typically higher level courses in your area of expertise (e.g., 400 level course on Software Testing). It is rare that you have to teach 100 level courses (e.g., Intro to Programming) because there are just too many students in these courses.

Most depts. would allow faculty to pick their favorite courses to teach.  In some cases, they might ask you to teach an existing core undergraduate courses (e.g., Theory, Operating Systems) and develop your own grad course (e.g., Software Verification).
It is likely that you will repeat the same courses. This significant reduces the amount of time you need to prepare for the course.  However, you should not teach the same course every semester, and you will want to alternate between grad and undergrad courses.

\begin{commentbox}[Vu]

    At GMU, I teach the same Software Specification and Construction course at both grad (SWE-619) and undergrad (SWE-419) levels.  While it is a bit boring to teach the same materials every semester, it is also nice because I don't have to prepare much for the course.  I did try to teach an AI Assurance seminar course once, but never liked it much because it only has MS students, who were not very interested in research.

    In contrast, I teach various courses at UNL. I teach the several undergraduate courses including Theory and design my own two courses on Compilers and Software Analysis. I also create a grad seminar on Software Verification. 
  \end{commentbox}

>>>>

While teaching takes about 40\% of your time, it is not as nearly as important as research for tenure.  Moreover, most new faculty do not prioritize teaching, e.g., compared to writing grants or recruiting students for research.  However, teaching is still important for your tenure case, and being a good teacher can be very rewarding. Innovative integration of research and teaching can also help you get good students and even NSF grants.

\section{Teaching Load}
New TT faculty often have a lighter teaching load in their first several years or until they get tenure.  This is to give you time to establish your research program.

At most large R1 universities the teaching load for TT faculty is 1+1 (1 course / semester). While some univs might give you a lighter load in your first year, most univs expect you to teach 1+1 starting your second year. In many cases, 1+1 is only for the first 2 or 3 years and after that it increases to 1+2 (or 2+1). It is rare that a TT in CS at R1 universities would have to teach 2+2 during the tenure track.

In some cases your contract might list 1+1 for the first X year and 1+2 after until tenure, but the dept unofficial gives TT faculty 1+1 until tenure. Some other ways to reduce teaching load include \emph{buying out} of teaching with a grant, leading a very large grant, and teaching many large classes.

\begin{commentbox}[Vu]
    At UNL, TT faculty is 1+1 for the first 3 years, and 1+2 after that.  At GMU, it is 1+1 for the first 3 years, but most TT faculty would continue this teaching load until they get tenure. In addition, GMU grants all TT faculty a semester off from teaching before tenure. Some faculty use this to work on their NSF CAREER proposal while others use it to prepare their tenure application. 
  \end{commentbox}

\section{Courses}

In most cases you will teach both undergrad and grad courses. New faculty often create graduate seminars and special topic courses to recruit students for their research group. Undergrad courses typically include higher level courses in your area of expertise (e.g., 400 level course on Software Testing). It is rare that you have to teach 100 level courses (e.g., Intro to Programming) because they are very large and thus time-consuming.

Most depts. would allow faculty to pick their favorite courses to teach.  In some cases, they might ask you to teach an existing core undergraduate courses (e.g., Theory, Operating Systems) and develop your own grad course (e.g., Software Verification).
It is likely that you get to repeat the same courses. This significant reduces the amount of time you need to prepare for the course.  However, you should not teach the same course every semester, and you will want to alternate between grad and undergrad courses.

\begin{commentbox}[Vu]
    At GMU, I teach the same Software Specification and Construction course at both grad (SWE-619) and undergrad (SWE-419) levels. This is very nice because I don't have to prepare much for the course.  I did try to teach an AI Assurance graduate seminar once, but never liked it much because it only has MS students, who were not very interested in research.

    In contrast, I teach various courses at UNL. I teach the several undergraduate courses including Theory, Software Engineering, and design my own two courses on Compilers and Software Analysis. I also create a grad seminar on Software Verification.

    I feel that UNL is a bit more flexible in creating new classes compared to GMU, where I am pretty much the main instructor for SWE-619/SWE-419.
  \end{commentbox}
